\documentclass{note}
\title{Quotes}
\date{2026-06-03}
\modified{2026-06-03}

\begin{document}

\blockquote{
  In my vision the computer acts as a transitional object
  to mediate relationships that are ultimately between person and person.
}{\bibref{mindstorms}}

\blockquote{
  Programs must be written for people to read,
  and only incidentally for machines to execute.
}{\bibref{sicp}}

\blockquote{
  Human beings learn by doing and by thinking about what they have done.
}{\bibref{pierce-information-theory}}

\blockquote{
  Great painters are often less skillful than mediocre painters;
  it is their concept of painting---not their skills---that defines their activity.
}{\bibref{clear-and-simple}}

\blockquote{
  You get no competitive edge from consuming the same stuff everyone else is consuming.
}{\bibref{hell-yeah-or-no}}

\blockquote{
  Hit songwriters often admit that their most successful hit song was one they thought was just stupid,
  even not worth recording.
  We're clearly bad judges of our own creations.
  We should just put them out there and let the world decide.
}{\bibref{hell-yeah-or-no}}

\blockquote{
  In Wrzesniewski's research,
  the happiest, most passionate employees
  are not those who followed their passion into a position,
  but instead those who have been around long enough
  to become good at what they do.
}{\bibref{newport-so-good}}

\blockquote{
  When you take a job
  take a long look at the people
  you're going to be working with---because
  the odds are
  you're going to become like them;
  they are not going to become like you.
  You can't change them.
  If it doesn't fit who you are,
  it's not going to work.
}{\bibref{barker-barking-up}}

\blockquote{
  There's nothing more effective for adopting new behaviors
  than surrounding yourself with people who already have them.
}{\bibref{forte-second-brain}}

\blockquote{
  An ideal world is left as an exercise to the reader.
}{\bibref[p. 109]{on-lisp}}

\blockquote{
  You can't go by what a girl says,
  when she's giving you the devil for making a chump of yourself.
  It's like Shakespeare.
  Sounds well, but doesn't mean anything.
}{\bibref{jeeves-joy-in-the-morning}}

\blockquote{
  Arrakis teaches the attitude of the knife---chopping off
  what's incomplete and saying:
  ``Now, it's complete because it's ended here.''
}{\bibref{dune}}

\blockquote{
  The urge to excel and the urge to lead aren’t the same.
  Sometimes I think they may be opposites.
}{\bibref{red-mars}}

\end{document}
